Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-10-17 writeups/dissertation/dissertation.tex
foo

@@

Vector-based models can be thought of as a compressed representation of the full co-occurrence matrix used for the ACT-R declarative memory system.
Each model computes an activation for each word to retrieve, given context and (for ACT-R's Bayesian model) prior knowledge.
Since both models produce activations for each chunk, one can be swapped out for another and used as the declarative memory component of the ACT-R infrastructure.
For example, \textcite{Rutledge2007} used a variant of the BEAGLE vector-based model as the DM component of ACT-R.
\citeAuthor{Rutledge2008} showed that an ACT-R theory with a vector-based memory system can successfully model the fan effect.

@@

Several hashtag recommendation models for Twitter have been developed recently.
One of the first models used a Bayesian co-occurrence statistical technique to predict the most likely hashtag associated with a post as a function of prior hashtag use and context \parencite{Mazzia2009}.
The model presented by \textcite{Mazzia2009} is quite similar to an ACT-R declarative retrieval model.
The main potential difference is that --- like \textcite{Kuo2011} and \textcite{Stanley2013} --- the global prior likelihood of a hashtag is computed without taking into account the specific user's past history.
Instead, the global prior is an overall average frequency of hashtag use across all users.
Further, no recency information was used to compute the global prior, so hashtags used in the recent past were not weighted more heavily. 
This was most likely not much of an issue for \citeAuthor{Mazzia2009}'s study, since all hashtags were collected within a relatively short period of time (about 3 months).

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
\textcite{Stanley2013} weighted each term in the total activation equation Table \ref{tabModACTRModel}) to maximize model accuracy.
\citeAuthor{Stanley2013} found that the optimal weights for the StackOverflow model for the prior term, title context term, and body context term were \num{.62}, \num{.93}, and \num{1.75} respectively.

@@

The runs from the Twitter popular-hashtags and StackOverflow randomly-sampled datasets were used for the analysis.
Both the random permutation and Bayesian models were tested.
Activations for context were computed by using five different methods to handle stop words:
The two removal methods used in \textcite{Sahlgren2008} (i.e., removal based on frequency, removal based on the pre-determined 571-word Cornell SMART stoplist),
the entropy-weighting method used in \textcite{Stanley2013},
combining the entropy-weighting and frequency-filtering methods,
and no removal or weighting of stop words.

@@

The ideal cutoff was determined by choosing the value where the standard deviation of the counts for each word for each of the tags were all relatively small.
The plots for the standard deviations across tags that were used to identify the cutoff for StackOverflow and Twitter are included in Figures \ref{figContextCutoffSO} and \ref{figContextCutoffT}.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are included in Figure \ref{figContextStopRPSO}.
The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\numNoZero{.45} to \numNoZero{.49}), while the entropy (\numNoZero{.60}) and frequency (\numNoZero{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-weighting technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

@@

The results comparing the full model to a context-only only model for the StackOverflow popular-users dataset are included in Figure \ref{figPUAddPrior}.
In this case, past tag history does contain unique variance for accurately predicting tag use.
Accuracy improves for both the Bayesian (\numNoZero{.51} to \numNoZero{.63}) and random permutation (\numNoZero{.47} to \numNoZero{.58}) models
when the prior model component is combined with context compared to context alone.

@@

This raises the question about how much past user behavior is enough for a site like StackOverflow to benefit from adding past user behavior as an additional predictor.
When the model predicts tags from a random sample of StackOverflow users, adding the prior component to the model does not significantly increase model accuracy compared to using context alone.
However, when the model predicts tags from a set of users that ask the most questions on the site, adding the prior does significantly increase accuracy.
For the StackOverflow dataset, a user asks on average 5 questions, and adds 3 tags to a post.
So the model uses approximately 15 prior observations when computing the prior component from a randomly-sampled on the site.
For the popular-users dataset used to generate the results in Figure \ref{figPUAddPrior}, a user asked 180 questions on average.
So for this dataset, the model uses approximately 540 prior observations when computing the prior activation.
So for the StackOverflow site, the prior component begins to generate unique predictive variance above and beyond the context component somewhere between when a user has generated 15 and 540 prior observations.
This result will be different for other sites that have collected a different amount of contextual information.
For sites like StackOverflow where the contextual component is already a strong predictor, it will take a larger number of prior observations before adding the prior component becomes beneficial.
Alternatively for sites like Twitter where contextual information is much more noisy and not as strong of a predictor, it will take fewer observations before adding the prior becomes beneficial.

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-userSjiSO-crop.pdf}}}
  \caption{Model accuracy with user-customized context component for StackOverflow}
  \label{figPUCustomSO}
\end{figure}

@@

\begin{table}[!ht]
  \caption{Summary of accuracy for each model and dataset}
  \label{tabOverallResults}
  {\tabulinesep=1.2mm
    \begin{tabu}{llll}
      \hline
	Site 		& Dataset		& Model						& Accuracy \\
	\hline
	StackOverflow	& popular-users		& Bayesian with entropy				& \numNoZero{.63} \\
			& 			& RP with entropy and log odds 			& \numNoZero{.58} \\
			& randomly-sampled	& Bayesian with entropy 			& \numNoZero{.63} \\
			& 			& RP with entropy and log odds			& \numNoZero{.63} \\
	Twitter		& popular-users		& Bayesian with entropy and user sji		& \numNoZero{.39} \\
			& 			& RP with entropy, log odds, and user sji	& \numNoZero{.39} \\
			& popular-hashtags	& Bayesian with entropy				& \numNoZero{.32} \\
			& 			& RP with entropy and log odds			& \numNoZero{.25} \\
	\hline
    \end{tabu}
  }
\end{table}

Overall, both the random permutation and Bayesian declarative memory models fit the observed tagging behavior in StackOverflow and Twitter relatively well.
The fits for StackOverflow in particular are quite high, especially given the large tag space on the site.
These results provide support for the idea that choosing a hashtag when composing a tweet and tagging a StackOverflow post is a declarative memory retrieval process.
However, both models were not 100 percent accurate, and the amount of variance left to predict was much larger than what would be observed if retrieval noise was included in the model.
So it is most likely the case that the task is not done solely with declarative memory processes.

Alternatively, perhaps authors are taking advantage of context information that is not currently represented in the model.
For example, time of day might be a useful predictor of tag use on Twitter.
\textcite{Cohn2012} built a tag-prediction model for Twitter that was based primarily on temporal cycles of hashtag use.
This technique could be naturally incorporated into a co-occurrence framework by including time of day as an additional context by tag co-occurrence for each post.
This would allow time of day information to be used alongside the rest of the context information in the words of the post.
A similar technique could be used to add additional context predictors to the model.
These co-occurrences could either be added on as additional columns to the existing co-occurrence matrix,
or a separate co-occurrence matrix could be used for each additional context component, similar to how separate matrices and context predictors were created for the words in the title and body of a post. 
Including additional context predictors such as time of day may help improve model accuracy,
and perhaps it is the case that the rest of the variance left to predict is from these context components that the author is using but are not yet included in the model.

@@

And compression for the random permutation model is essentially random.
That is, no SVD is done to identify the most predictive components,
so the compressed representation used by the random permutation model does not contain nearly as much of the variance from the original co-occurrence matrix that it could have.
However, performing a full SVD on a co-occurrence matrix as large as the ones generated for StackOverflow and Twitter is computationally infeasible.
So, it also seems plausible that this SVD computation is infeasible at the neural level.
When deriving the LSA approach, \textcite{Landauer1997} explicitly state that they make no claims that the brain is computing an SVD at the neural level;
only that the neural machinery is set up such that the representation makes predictions that are similar to what would have been generated if the matrix decomposition for an SVD would have been computed.
Perhaps the randomly-compressed representation used for the random permutation model is the way in which the brain tries to efficiently achieve the full LSA- and SVD-like approach that \citeauthor{Landauer1997} propose.

@@

The Bayesian declarative memory model for ACT-R has an attentional weight term $W_{j}$ that can be used to implement how stop words are handled.
The random permutation model was also easily modified to have an attentional weight term by attenuating each word's set of ones and negative ones values by that word's attentional weight.
Both a method for stop-word removal (frequency filtering) and stop-word attenuating (entropy weighting) performed well for both models.
Also, using a proper method for handling the low-predictor words produced some of the largest improvements in model accuracy compared to the other explorations
(e.g., compared to word order, properly adding terms, compression amount for random permutation representation).
Since the ACT-R declarative memory theory already has a component for attentional weight, it may be the case that the neural system performs some form of stop-word weighting when computing activations,
and this is reflected in the attentional weight term.
That is, $W_{j}$ can be defined mathematically, has the same general form across tasks, and both the frequency-filtering and entropy-weighting methods for stop-word removal are likely first candidates for the definition.
This claim, however, would require that no top-down activation is used to compute the attentional weight, which would mean that the $W_{j}$ is not under conscious control, which is not likely to be correct.
Perhaps it is a combination of top-down and bottom-up systems that determine $W_{j}$, and the entropy-weighting and frequency-filtering techniques are methods of formalizing the bottom-up component.

@@

The Bayesian and vector-based memory models differ in how the co-occurrence matrix is represented (lossy compression for vector-based models).
They also differ in how the activation of each tag is calculated, given the different representations.
The vector-based memory models, like LSA, compute a correlation between the context and each tag, while the Bayesian models sum the log likelihoods that each word in the context occurs with each tag.
These two methods for computing the activation are correlated but not identical.
Further research could try to tease apart how model accuracy is independently influenced by the representation (compressed or not) and the activation calculation (correlation or log likelihood).
To do this, four models could be constructed (uncompressed and compressed representation crossed with correlation and log likelihood activation) compared to the two in this study,
so that the effects of representation and activation calculation could be partialed out.
This would allow, for example, the method of compression for the random permutation vector-based model to be analyzed independent of activation calculation.
The results would provide additional evidence to help determine the specific representation that the declarative memory system uses to represent co-occurrences,
and also the specific activation calculation that is carried out by the system.

@@

The tag-recommendation models developed through this research should be immediately relevant to real-world tasks.
Model accuracy is high enough (particularly for StackOverflow) for the developed models to be used as a foundation for tag recommendation systems that are deployed on Twitter or StackOverflow.
Also, since the developed models are task agnostic, they can potentially be used as tag recommendation systems for other domains.
Examples of similar tagging tasks are hashtag use on Facebook, flagging (i.e., tagging), Wikipedia articles for content issues or overly opinionated material,
auto labeling incoming mail and placing it into appropriate folders, and spam filtering.
Having an accurate and efficient co-occurrence memory model that correctly retrieves relevant tags can be applied to a wide variety of tasks.

>>>>

% Changing horizontal space between subsubsection title and text, without loading titlesec package.
% Not loading titlesec because it clashes with apa6 package.
% Took base \@startsection command from apa6 .cls file, and tweaked horizontal spacing. 
% http://tex.stackexchange.com/questions/22576/redefining-sectioning-commands
% http://www.tex.ac.uk/cgi-bin/texfaq2html?label=atsigns
% http://tex.stackexchange.com/questions/60127/subsubsection-description-and-title-not-separated
\makeatletter
\renewcommand{\subsubsection}{%
  \@startsection
  {subsubsection}%
  {3}%
  {\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-.4em}%
  {\normalfont\normalsize\bfseries\addperi}}
\renewcommand{\paragraph}{%
  \@startsection
  {paragraph}%
  {4}%
  {\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-.4em}%
  {\normalfont\normalsize\bfseries\itshape\addperi}}
\makeatother

@@

Vector-based models can be thought of as a compressed representation of the full co-occurrence matrix used for the ACT-R declarative memory system.
Each model computes an activation for each word to retrieve, given context and (for ACT-R's Bayesian model) prior knowledge.
Since both models produce activations for each chunk, one can be swapped out for another and used as the declarative memory component of the ACT-R infrastructure.
For example, \textcite{Rutledge2007} used a variant of the BEAGLE vector-based model as the DM component of ACT-R.
\citeauthor{Rutledge2008} showed that an ACT-R theory with a vector-based memory system can successfully model the fan effect.

@@

Several hashtag recommendation models for Twitter have been developed recently.
One of the first models used a Bayesian co-occurrence statistical technique to predict the most likely hashtag associated with a post as a function of prior hashtag use and context \parencite{Mazzia2009}.
The model presented by \textcite{Mazzia2009} is quite similar to an ACT-R declarative retrieval model.
The main potential difference is that --- like \textcite{Kuo2011} and \textcite{Stanley2013} --- the global prior likelihood of a hashtag is computed without taking into account the specific user's past history.
Instead, the global prior is an overall average frequency of hashtag use across all users.
Further, no recency information was used to compute the global prior, so hashtags used in the recent past were not weighted more heavily. 
This was most likely not much of an issue for \citeauthor{Mazzia2009}'s study, since all hashtags were collected within a relatively short period of time (about 3 months).

@@

In order to best combine these predictors, a logistic regression statistical technique similar to \textcite{Stanley2013} was used to determine the most optimal weights for each term.
\textcite{Stanley2013} weighted each term in the total activation equation Table \ref{tabModACTRModel}) to maximize model accuracy.
\citeauthor{Stanley2013} found that the optimal weights for the StackOverflow model for the prior term, title context term, and body context term were \num{.62}, \num{.93}, and \num{1.75} respectively.

@@

The runs from the Twitter popular-hashtags and StackOverflow randomly-sampled datasets were used for the analysis.
Both the random permutation and Bayesian models were tested.
Activations for context were computed by using five different methods to handle stop words:
The two removal methods used in \textcite{Sahlgren2008}
(i.e., removal based on frequency, removal based on the pre-determined 571-word Cornell SMART stoplist \parencite{Salton1988}),
the entropy-weighting method used in \textcite{Stanley2013},
combining the entropy-weighting and frequency-filtering methods,
and no removal or weighting of stop words.

@@

The ideal cutoff was determined by choosing the value where the standard deviation of the counts for each word for each of the tags were all relatively small.
The plots for the standard deviations across tags that were used to identify the cutoff for StackOverflow and Twitter are shown in Figures \ref{figContextCutoffSO} and \ref{figContextCutoffT}.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are shown in Figure \ref{figContextStopRPSO}.
The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\numNoZero{.45} to \numNoZero{.49}), while the entropy (\numNoZero{.60}) and frequency (\numNoZero{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-weighting technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

@@

The results comparing the full model to a context-only only model for the StackOverflow popular-users dataset are depicted in Figure \ref{figPUAddPrior}.
In this case, past tag history does contain unique variance for accurately predicting tag use.
Accuracy improves for both the Bayesian (\numNoZero{.51} to \numNoZero{.63}) and random permutation (\numNoZero{.47} to \numNoZero{.58}) models
when the prior model component is combined with context compared to context alone.

@@

This raises an important applied question: Exactly how much past user behavior is enough for a site like StackOverflow to benefit from adding past user behavior as an additional predictor.
When the model predicts tags from a random sample of StackOverflow users, adding the prior component to the model does not significantly increase model accuracy compared to using context alone.
However, when the model predicts tags from a set of users that ask the most questions on the site, adding the prior does significantly increase accuracy.
For the StackOverflow dataset, a user asks on average 5 questions and adds 3 tags to a post.
So the model uses approximately 15 prior observations when computing the prior component for a randomly-sampled user on the site.
For the popular-users dataset used to generate the results in Figure \ref{figPUAddPrior}, a user asked 180 questions on average.
So for this dataset, the model uses approximately 540 prior observations when computing the prior activation.
So for the StackOverflow site, the prior component begins to generate unique predictive variance above and beyond the context component somewhere between when a user has generated 15 and 540 prior observations.
This result will be different for other sites that have collected a different amount of contextual information.
For sites like StackOverflow where the contextual component is already a strong predictor, it will take a larger number of prior observations before adding the prior component becomes beneficial.
Alternatively for sites like Twitter where contextual information is much more noisy and not as strong of a predictor, it will take fewer observations before adding the prior becomes beneficial.

@@

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-userSjiSO-crop.pdf}}}
  \caption{
    Model accuracy with user-customized context component for StackOverflow.
    A model name with ``user sji'' uses a co-occurrence matrix customized for each user instead of a single global co-occurrence matrix.
  }
  \label{figPUCustomSO}
\end{figure}

@@

\begin{table}[!ht]
  \caption{
    Summary of accuracy for each model and dataset.
    The Bayesian model includes entropy weighting for the context words,
    the RP model includes entropy weighting and log odds transformation when combining the context and prior components.
    Both models use a user-customized co-occurrence matrix for the Twitter popular-users dataset, since a global co-occurrence
    matrix is not available for this dataset.
  }
  \label{tabOverallResults}
  {\tabulinesep=1.2mm
    \begin{tabu}{llll}
      \hline
	Site 		& Dataset		& Model						& Accuracy \\
	\hline
	StackOverflow	& popular-users		& Bayesian with entropy				& \numNoZero{.63} \\
			& 			& RP with entropy and log odds 			& \numNoZero{.58} \\
			& randomly-sampled	& Bayesian with entropy 			& \numNoZero{.63} \\
			& 			& RP with entropy and log odds			& \numNoZero{.63} \\
	Twitter		& popular-users		& Bayesian with entropy and user sji		& \numNoZero{.39} \\
			& 			& RP with entropy, log odds, and user sji	& \numNoZero{.39} \\
			& popular-hashtags	& Bayesian with entropy				& \numNoZero{.32} \\
			& 			& RP with entropy and log odds			& \numNoZero{.25} \\
	\hline
    \end{tabu}
  }
\end{table}

Overall, both the random permutation and Bayesian declarative memory models fit the observed tagging behavior in StackOverflow and Twitter relatively well.
The fits for StackOverflow in particular are quite high, especially given the large tag space on the site.
These results provide support for the idea that choosing a hashtag when composing a tweet and tagging a StackOverflow post is akin to a declarative memory retrieval process.
However, both models were not 100 percent accurate, and the amount of variance left to predict was much larger than what would be observed if retrieval noise was included in the model.
So it is most likely the case that the task is not done solely with declarative memory processes.

Alternatively, perhaps authors of posts are taking advantage of context information that is not currently represented in the model.
For example, time of day might be a useful predictor of tag use on Twitter.
\textcite{Cohn2012} built a tag-prediction model for Twitter that was based primarily on temporal cycles of hashtag use.
This technique could be naturally incorporated into a co-occurrence framework by including time of day as an additional context by tag co-occurrence for each post.
This would allow time of day information to be used alongside the rest of the context information in the words of the post.
A similar technique could be used to add additional context predictors to the model.
These co-occurrences could either be added on as additional columns to the existing co-occurrence matrix,
or a separate co-occurrence matrix could be used for each additional context component, similar to how separate matrices and context predictors were created for the words in the title and body of a post. 
Including additional context predictors such as time of day may help improve model accuracy,
and perhaps it is the case that the rest of the variance left to predict is from these context components that the author is using but are not yet included in the model.

@@

As previously noted, compression for the random permutation model is essentially random.
That is, no SVD is done to identify the most predictive components,
so the compressed representation used by the random permutation model does not contain nearly as much of the variance from the original co-occurrence matrix that it could have.
However, performing a full SVD on a co-occurrence matrix as large as the ones generated for StackOverflow and Twitter is computationally infeasible.
Also, when deriving the LSA approach, \textcite{Landauer1997} explicitly state that they make no claims that the brain is computing an SVD at the neural level;
only that the neural machinery is set up such that the representation makes predictions that are similar to what would have been generated if the matrix decomposition for an SVD would have been computed.
Perhaps the randomly-compressed representation used for the random permutation model is the way in which the brain tries to efficiently achieve the full LSA- and SVD-like approach that \citeauthor{Landauer1997} propose.

@@

The Bayesian declarative memory model for ACT-R has an attentional weight term $W_{j}$ that can be used to attenuate or remove stop words.
The random permutation model was also easily modified to have an attentional weight term by attenuating each word's set of ones and negative ones values by that word's attentional weight.
Both a method for stop-word removal (frequency filtering) and stop-word attenuating (entropy weighting) performed well for both models.
Also, using a proper method for handling the low-predictor words produced some of the largest improvements in model accuracy compared to the other explorations
(e.g., compared to word order, properly adding terms, compression amount for random permutation representation).
Since the ACT-R declarative memory theory already has a component for attentional weight, it may be the case that the neural system performs some form of stop-word weighting when computing activations,
and this is reflected in the attentional weight term.
That is, $W_{j}$ can be defined mathematically, has the same general form across tasks, and both the frequency-filtering and entropy-weighting methods for stop-word removal are likely first candidates for the definition.
This claim, however, would require that no top-down activation is used to compute the attentional weight, which would mean that the $W_{j}$ is not under conscious control, which is not likely to be correct.
Perhaps it is a combination of top-down and bottom-up systems that determine $W_{j}$, and the entropy-weighting and frequency-filtering techniques are methods of formalizing the bottom-up component.

@@

The Bayesian and vector-based memory models differ in how the co-occurrence matrix is represented (lossy compression for vector-based models).
They also differ in how the activation of each tag is calculated, given the different representations.
The vector-based memory models, like LSA, compute a correlation between the context and each tag, while the Bayesian models sum the log likelihoods that each word in the context occurs with each tag.
These two methods for computing the activation are correlated but not identical.
Further research could try to tease apart how model accuracy is independently influenced by the representation (compressed or not) and the activation calculation (correlation or log likelihood).
To do this, four models could be constructed (uncompressed and compressed representation crossed with correlation and log likelihood activation) compared to the two in this study,
so that the effects of representation and activation calculation could be partialed out.
This would allow, for example, the method of compression for the random permutation vector-based model to be analyzed independent of activation calculation.
The results would help determine the specific representation that the declarative memory system uses to represent co-occurrences,
and also the specific activation calculation that is carried out by the system.

@@

The tag-recommendation models developed through this research should be relevant to real-world tasks.
Model accuracy is high enough (particularly for StackOverflow) for the developed models to be used as a foundation for tag recommendation systems that are deployed on Twitter or StackOverflow.
However, the focus for this research was to evaluate the accuracy of two cognitively-plausible declarative memory models, so the models were more constrained than general machine learning models of tagging.
So it is unlikely that these general models of declarative memory will outperform specific machine learning models that have been optimized for a specific tagging task. 
Nevertheless, one of the advantages of the general declarative memory models is that they are task agnostic, so it would take minimal time to deploy these models on a new tagging task.
